The Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-08 ms.tex
2.2 update

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the density $\rho$, temperature $T$, and radial velocity $v$ of the CNM for each galaxy model by solving the equations of one-dimensional, time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\rm enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] 
, 
\label{eq:dsdt}
\end{align}
where $p$ and $s$ are the pressure and specific entropy, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p = \rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$.

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a star burst (e.g., \citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the overall mass return rate from evolved stars as a function of B-band luminosity.  Although we give the mass return rate per unit volume, our expressions are equivalent.}, although we will show later that $\eta$ can be significantly higher if star formation is ongoing (Fig.~\ref{fig:eta}).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because isotropic
mass injection in the SMBH rest frame represents a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected locally
by stellar winds and the mean properties of the background flow.  The
term $\propto -q c_{s}^{2}$ is important because it acts to potentially stabilize the flow to runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random movement
of stars with respect to one another at the characteristic velocity
$\sim \sigma \approx \sqrt{GM_{\rm enc}/r}$.  This heating is present even in old stellar populations for which mass and energy injection are dominated by AGB winds with low velocities $\ll \sigma$.  The second term, $v_{w}^{2}$, parameterizes
the contribution of additional sources of heating, such as faster main
sequence or Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars, supernovae, AGN
feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ is uncertain and will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Heat transport by electron conductivity results in an an additional term on the right hand side of the entropy equation of the form  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering becomes comparable to the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field oriented perpendicular to the flow could in principle suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).

Radiative cooling is also not included in our baseline calculations.  Across much of the parameter space radiative cooling is subdominant compared to other sources of cooling.  Radiative cooling becomes important when $\vwO$ is low and/or $\eta$ is high, but once important it leads to thermal instabilities (e.g.~\citealt{McCourt+12}) which our 1D
time-independent model cannot account for self-consistently anyways.  Indeed, one of the main goals of this work is to use our analytic and numerical solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities will develop ($\S\ref{sec:cooling}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the
steady-state inflow/outflow structure (assuming it exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.

@@

\subsection{Importance of Heat Conduction}

>>>>

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the density $\rho$, temperature $T$, and radial velocity $v$ of the CNM for each galaxy model by solving the equations of one-dimensional, time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\rm enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] 
, 
\label{eq:dsdt}
\end{align}
where $p$ and $s$ are the pressure and specific entropy, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We adopt an ideal gas equation of state with $p = \rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$.

The source term in equation (\ref{eq:drhodt}),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
yr.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following an impulsive star burst (e.g., \citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the mass return rate from evolved stars as a function of B-band luminosity instead of volumetrically, but our expressions are equivalent.}, although $\eta$ is significantly higher for continuous star formation histories (Fig.~\ref{fig:eta}).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because the isotropic injection of mass represents, in the SMBH rest frame, a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected locally
by stellar winds and the mean properties of the background flow.  The
term $\propto -q c_{s}^{2}$ is important because it acts to potentially stabilize the flow to runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random movement
of stars with respect to one another at the characteristic velocity of $\sigma \approx \sqrt{GM_{\rm enc}/r}$.  This heating is present even in old stellar populations for which mass and energy injection are dominated by AGB winds with low velocities $\ll \sigma$.  The second term, $v_{w}^{2}$, parameterizes all additional sources of heating, including faster main
sequence or Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars, supernovae, AGN
feedback, etc.  As will be discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Electron heat conduction results in an an additional term on the right hand side of the entropy equation of the form:  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering exceeds the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field that is oriented perpendicular to the flow could suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).  We show later in $\S\ref{sec:conductivity}$ that neglecting conductivity results in at most only order-unity corrections to the key properties of solutions calculated assuming its presence.

We also neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low and/or stellar mass loss rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instabilities (e.g.~\citealt{McCourt+12}) that our 1D
time-independent model cannot account for self-consistently anyways.  A main goal of this work is to use analytic and numerical solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:cooling}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the
steady-state inflow/outflow solution (assuming one exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.

@@

\subsection{Importance of Heat Conduction}
\label{sec:conductivity}
